New evidence of self-care tooling and phylogenetic modelling reveal parrot tool use is not rare

### 2025-02-05 appendix.Rmd
prepare for submission
update manuscript
update manuscript and prepare for submission
revise and resubmit

\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\tablename}{Supplementary Table}
\renewcommand{\thefigure}{\arabic{figure}} \setcounter{figure}{0}
\renewcommand{\thetable}{\arabic{table}} \setcounter{table}{0}
\renewcommand{\theequation}{S\arabic{equation}} \setcounter{equation}{0}

# **Supplementary Information**

\setcounter{page}{1}
\centering
\noindent \hspace*{5mm} \small Crowdsourcing and phylogenetic modelling reveal parrot tool use is not rare \newline
\hspace*{1cm} \small Amalia P. M. Bastos\textsuperscript{1,2}, Scott Claessens\textsuperscript{2}, Ximena J. Nelson\textsuperscript{3}, David Welch\textsuperscript{4}, \newline
\hspace*{25mm} Quentin D. Atkinson\textsuperscript{2}, Alex H. Taylor\textsuperscript{2,3,5,6} \newline
\raggedright
\noindent \footnotesize \textsuperscript{1} Department of Psychological & Brain Sciences, Johns Hopkins University, Baltimore, MD, United States \newline
\noindent \footnotesize \textsuperscript{2} School of Psychology, University of Auckland, Auckland, New Zealand \newline
\noindent \footnotesize \textsuperscript{3} School of Biological Sciences, University of Canterbury, Christchurch, New Zealand \newline
\noindent \footnotesize \textsuperscript{4} School of Computer Science, University of Auckland, Auckland, New Zealand \newline
\noindent \footnotesize \textsuperscript{5} ICREA, Pg. Lluís Companys 23, Barcelona, Spain \newline
\noindent \footnotesize \textsuperscript{6} Institute of Neuroscience, Universitat Autònoma de Barcelona, Barcelona, Spain \newline
\normalsize
\newpage

## Supplementary Figures

@@

```{r plotComparison, echo=FALSE, warning=FALSE, fig.width=8, fig.height=4, fig.cap="(ref:plotComparisonCaption)"}
tar_read(plotComparison)
```

@@

```{r plotASR1, echo=FALSE, warning=FALSE, fig.width=10, fig.height=13, fig.cap="(ref:plotASR1Caption)"}
tar_read(plotASR1)
```

@@

```{r plotASR2, echo=FALSE, warning=FALSE, fig.width=10, fig.height=13, fig.cap="(ref:plotASR2Caption)"}
tar_read(plotASR2)
```

@@

```{r plotASR3, echo=FALSE, warning=FALSE, fig.width=10, fig.height=13, fig.cap="(ref:plotASR3Caption)"}
tar_read(plotASR3)
```

>>>>

\renewcommand{\figurename}{Figure}
\renewcommand{\tablename}{Table}
\renewcommand{\thefigure}{S\arabic{figure}} \setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}} \setcounter{table}{0}
\renewcommand{\theequation}{S\arabic{equation}} \setcounter{equation}{0}

# **Supplemental Information**

\setcounter{page}{1}
\centering
\noindent \hspace*{5mm} \small New evidence of self-care tooling and phylogenetic modelling reveal parrot tool use is not rare \newline
\hspace*{1cm} \small Amalia P. M. Bastos\textsuperscript{1,2}, Scott Claessens\textsuperscript{2}, Ximena J. Nelson\textsuperscript{3}, David Welch\textsuperscript{4}, \newline
\hspace*{25mm} Quentin D. Atkinson\textsuperscript{2}, Alex H. Taylor\textsuperscript{2,3,5,6} \newline
\raggedright
\noindent \footnotesize \textsuperscript{1} Department of Psychological & Brain Sciences, Johns Hopkins University, Baltimore, MD, United States \newline
\noindent \footnotesize \textsuperscript{2} School of Psychology, University of Auckland, Auckland, New Zealand \newline
\noindent \footnotesize \textsuperscript{3} School of Biological Sciences, University of Canterbury, Christchurch, New Zealand \newline
\noindent \footnotesize \textsuperscript{4} School of Computer Science, University of Auckland, Auckland, New Zealand \newline
\noindent \footnotesize \textsuperscript{5} ICREA, Pg. Lluís Companys 23, Barcelona, Spain \newline
\noindent \footnotesize \textsuperscript{6} Institute of Neuroscience, Universitat Autònoma de Barcelona, Barcelona, Spain \newline
\normalsize
\newpage

## Supplemental Figures

@@

```{r plotComparison, echo=FALSE, warning=FALSE, fig.width=8, fig.height=4, fig.cap="(ref:plotComparisonCaption)"}
#tar_read(plotComparison)
knitr::include_graphics("figures/comparison.pdf")
```

@@

```{r plotASR1, echo=FALSE, warning=FALSE, fig.width=10, fig.height=13, fig.cap="(ref:plotASR1Caption)"}
#tar_read(plotASR1)
knitr::include_graphics("figures/asr1.pdf")
```

@@

```{r plotASR2, echo=FALSE, warning=FALSE, fig.width=10, fig.height=13, fig.cap="(ref:plotASR2Caption)"}
#tar_read(plotASR2)
knitr::include_graphics("figures/asr2.pdf")
```

@@

```{r plotASR3, echo=FALSE, warning=FALSE, fig.width=10, fig.height=13, fig.cap="(ref:plotASR3Caption)"}
#tar_read(plotASR3)
knitr::include_graphics("figures/asr3.pdf")
```

### 2025-02-05 manuscript.Rmd
revise and resubmit

Comparative work on the evolution of tool use is a paradigmatic example of this 
issue. The initial discoveries of tool use in chimpanzees [@Goodall1964], 
birds [@Hunt1996], dolphins [@Smolker1997], and octopuses [@Finn2009] occurred 
decades after significant advances on other more easily measurable aspects of 
their biology. Since then, scholars have proposed a clear operational 
definition of tool use applicable to all species --- the manipulation of an 
unattached object as an extension of the animal's body to achieve a 
goal [@Shumaker2011] --- and have used the distribution of species meeting this 
definition to make various claims about the evolutionary drivers of tool use 
behaviours. For example, based on the observation that bird species with 
reported tool use tend to have larger brains, researchers have identified 
higher relative brain size as a likely precondition for tool using 
capabilities [@Audet2024;@Lefebvre2002;@Lefebvre1997;@Lefebvre2004]. These 
researchers argue that larger brains are better able to integrate visual and 
somatosensory information when innovating novel behaviours, such as tool use, 
in changing environments [@Sol2005;@Sol2009] (but see ref [@Kaplan2020]). 
Similarly, researchers have used existing reports of tool use in birds to debate
the roles of generalist versus specialist feeding strategies in driving the 
evolution of tool use, with some arguing that feeding generalists require 
technical innovations to expand their dietary niche 
[@Ducatez2015;@Overington2011;@Sol2005] and others arguing that feeding 
specialists require technical innovations for extractive foraging of specific 
foods [@HenkeVonDerMalsburg2020;@MettkeHofmann2014].

However, before we can make claims of this kind, we need to know whether current 
research effort in the literature is sufficient for robust conclusions to be 
drawn about the evolution of tool use. In fact, evidence suggests that research 
effort is often systematically biased towards particular taxonomic groups, parts 
of the world that are easy to access, and species with life history traits that 
make them easier to study, such as larger distribution ranges and population 
sizes [@Ducatez2014]. This is a crucial limitation because insufficient 
observation may lead researchers to miss true instances of tool behaviours and 
thus draw premature conclusions about the evolutionary drivers and origins of 
tool use. Researchers are keenly aware of this problem and have attempted to 
deal with it in different ways, such as controlling for the number of scientific 
papers published on different species. But previous work has not yet attempted 
to quantify and explicitly model the relationship between actual tool-using 
behaviour and what is reported in the scientific literature. If more tool-using 
species exist than previously thought, this could have important implications 
for theories of the evolutionary drivers and origins of tool use and for our 
understanding of how rare this behaviour actually is.

@@

Figure \@ref(fig:plotPhylo1) maps the findings from the video survey onto a 
maximum clade credibility phylogeny for the parrot order, plotted alongside 
species previously identified in the scientific literature. Before the video 
survey, 11 parrot species (3%) had been identified as tool users in the 
scientific literature. Across our video survey, we observed true tool use for self-care in 
`r trueCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()` 
species, 5 of which overlapped with the scientific literature and 
`r (trueCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()) - 5` 
of which were novel species. All of the species identified in the video survey 
were cockatoos (_Cacatuidae_), Old World parrots (_Psittacinae_), or neotropical 
parrots (_Arinae_). The most common species in our survey, accounting for 
`r crowdData %>% filter(Species == "Pyrrhura molinae") %>% nrow()` videos from 
`r crowdData %>% filter(Species == "Pyrrhura molinae") %>% pull(IndividualID) %>% unique() %>% length()` 
individuals, was the green-cheeked conure (_Pyrrhura molinae_). In accordance 
with the scientific literature, the video survey found no evidence of tool use 
in any species of Psittaculidae, despite this family containing some of the most 
commonly kept pet species, including lovebirds, lorikeets, and Asian parakeet 
species. Combining both the video survey and the scientific literature, we can 
thus identify 28 total parrot species (7%) capable of true tool use as defined 
by ref [@Shumaker2011], compared to the 11 previously reported.

@@

Our data raise the question of why such a wide range of parrot species exhibit 
self-scratching tool use behaviours in captive environments. It is possible that 
these self-care behaviours emerge more often in captivity because pet parrots 
are kept with few or no conspecifics, and therefore must innovate tool use to 
preen inaccessible body areas, such as their heads and necks, which would have 
been otherwise preened by flock members. This may indicate that human 
interaction and allopreening in the absence of a conspecific is insufficient for
pet parrots, as suggested by studies on welfare and chronic stress in lone 
captive parrots [@Aydinonat2014;@Engebretson2006;@Mason2010;@Meehan2003]. Beyond 
social isolation in captivity, other factors that reduce preening efficacy by an 
individual might also lead to the innovation of self-care tooling: for example,
a captive kea housed with conspecifics in a naturalistic environment was shown
to have innovated the use of a pebble tool for self-care likely as a consequence 
of a missing upper mandible [@Bastos2021]. These behaviours might therefore be 
innovated by individuals when ecologically necessary. Critically, however, a 
capacity to innovate a behaviour such as tool use must exist in a species in 
order to manifest in any environment, whether that be captive or in the wild. 
The fact that only a subset of parrot species are innovating self-care tooling 
in response to their captive setting and that self-care tooling is predicted by 
phylogenetic history and encephalisation strongly suggests that self-scratching 
indicates a capacity to produce tool use, which is expressed in different ways 
in captive and wild settings.

Our findings have a number of important implications. First, our findings show 
that current research effort in the scientific literature is insufficient to 
capture the real world occurrence of parrot tool use. If the scientific 
literature had sampled the natural world sufficiently, we would expect to see 
close correspondence between those species reported as tool users in the 
literature and those species the public have filmed performing tool use. 
Instead, we discovered a large discrepancy between these two data sources, both 
in the prevalence of tool use and the species identified, likely due to the 
numerous difficulties inherent to conducting observations in the wild. This 
raises the possibility that other rare behaviours remain yet to be discovered in 
the scientific literature.

@@

Third, our results support existing theories of the drivers of tool use. We 
found that encephalisation was strongly positively related to the probability of 
tool use in our phylogenetic model, supporting previous theories linking 
relative brain size to increased tool innovation in birds [@Lefebvre1997;
@Lefebvre2002;@Lefebvre2004] and primates [@Reader2002]. We acknowledge that
encephalisation quotient is not a perfect measure due to measurement error and
challenges with interpretation [@Healy2007;@Logan2018]. However, encephalisation
quotient has much greater coverage across the parrot phylogeny than more 
fine-grained measures like whole neuron count [@Sol2022] and there is no reason 
to think that measurement error would produce the consistent patterns across our
study and prior work. To understand the causal mechanisms responsible for these
relationships, we encourage further work on the specific neural correlates of 
technical intelligence in parrots, e.g., [@Alvarez2020]. In our phylogenetic 
model, we also found that tool use was somewhat more likely among feeding 
generalists compared to feeding specialists, although this difference was 
uncertain. This trend supports previous suggestions that increased cognitive 
abilities and technical innovation rates are required to expand a generalist 
species' dietary niche [@HenkeVonDerMalsburg2020;@HenkeVonDerMalsburg2021;
@Ducatez2015]. However, the trend contradicts theories linking tool use to 
dietary specialisation, whereby species eating specific foods that require 
extractive foraging have higher cognitive ability and are especially prone to 
using tools [@HenkeVonDerMalsburg2020].

A potential concern with data originating from digital video platforms is its 
reliability [@Auersperg2020;@DechaumeMoncharmont2020;@Farrar2020;@Nelson2013;@Sandor2020;@vonBayern2020]. 
In particular, the self-care tool use instances detected in our video survey 
might be attributable to training or unintentional cueing by the birds' owners.
However, there was little evidence to suggest that the observations of 
self-scratching tool use in the video survey were merely unintentional accidents 
or explicitly trained behaviours. Individual parrots often used tools slowly and 
repetitively over long periods of time, even across multiple different videos, 
suggesting that their behaviour was not random or accidental [@Sandor2020]. 
Parrots preferentially employed self-scratching tools on areas of their body 
that were otherwise inaccessible, with 96% of all instances involving scratching 
of the head or neck, suggesting intentional tool use. For 60% of the species in 
the video survey, we found two or more videos of repetitive and sustained 
scratching by different individuals of the same species in separate households, 
suggesting that the manipulations were intentional and recurring events that did 
not represent unusual stereotypies of any single individual. Regarding the 
possibility of training or cueing from owners, over half of the videos contained 
no evidence of human interaction aside from filming the behaviour. Humans only 
handed parrots their tools in two of the videos, and none of the videos featured 
owners directly rewarding tool use behaviours with food. It is unlikely that 
owners would specifically train parrots to scratch themselves with tools. 
Finally, the high levels of phylogenetic signal in our data provide strong 
evidence that the observations from our video survey reflect
biologically-endowed capacities for tool use rather than accidental or trained
behaviours, which would likely appear uniformly across the phylogeny. 
Nevertheless, future work using data such as this should choose target 
behaviours carefully and examine videos for evidence of training or cueing.

@@

@@

This project was made possible through the support of a grant from the Templeton 
World Charity Foundation (A.H.T., X.J.N.). The authors would like to thank 
Daniel Sol for providing feedback on a previous version of the manuscript.

>>>>

Comparative work on the evolution of tool use is a paradigmatic example of this 
issue. The initial discoveries of tool use in chimpanzees [@Goodall1964], 
birds [@Hunt1996], dolphins [@Smolker1997], and octopuses [@Finn2009] occurred 
decades after significant advances on other more easily measurable aspects of 
their biology. Since then, scholars have proposed a clear operational 
definition of tool use applicable to all species --- the manipulation of an 
unattached object as an extension of the animal's body to achieve a 
goal [@Shumaker2011] --- and have used the distribution of species meeting this 
definition to make various claims about the evolutionary drivers of tool use 
behaviours. For example, based on the observation that bird species with 
reported tool use tend to have larger brains, researchers have identified 
higher relative brain size as a likely precondition for tool using 
capabilities [@Audet2024;@Lefebvre2002;@Lefebvre1997;@Lefebvre2004]. These 
researchers argue that larger brains are better able to integrate visual and 
somatosensory information when innovating novel behaviours, such as tool use, 
in changing environments [@Sol2005;@Sol2009] (but see Kaplan[@Kaplan2020]). 
Similarly, researchers have used existing reports of tool use in birds to debate
the roles of generalist versus specialist feeding strategies in driving the 
evolution of tool use, with some arguing that feeding generalists require 
technical innovations to expand their dietary niche 
[@Ducatez2015;@Overington2011;@Sol2005] and others arguing that feeding 
specialists require technical innovations for extractive foraging of specific 
foods [@HenkeVonDerMalsburg2020;@MettkeHofmann2014].

However, before we can make claims of this kind, we need to know whether current 
research effort in the literature is sufficient for robust conclusions to be 
drawn about the evolution of tool use. In fact, evidence suggests that research 
effort is often systematically biased towards particular taxonomic groups, parts 
of the world that are easy to access, and species with life history traits that 
make them easier to study, such as larger distribution ranges and population 
sizes [@Ducatez2014]. This is a crucial limitation because insufficient 
observation may lead researchers to miss true instances of tool behaviours and 
thus draw premature conclusions about the evolutionary drivers and origins of 
tool use. Researchers are keenly aware of this problem and have attempted to 
deal with it in different ways, such as controlling for the number of scientific 
papers published on different species. However, previous work has not yet 
attempted to quantify and explicitly model the relationship between actual
tool-using behaviour and what is reported in the scientific literature. If more
tool-using species exist than previously thought, this could have important
implications for theories of the evolutionary drivers and origins of tool use
and for our understanding of how rare this behaviour actually is.

@@

Figure \@ref(fig:plotPhylo1) maps the findings from the video survey onto a 
maximum clade credibility phylogeny for the parrot order, plotted alongside 
species previously identified in the scientific literature. Before the video 
survey, 11 parrot species (3%) had been identified as tool users in the 
scientific literature. Across our video survey, we observed true tool use for self-care in 
`r trueCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()` 
species, 5 of which overlapped with the scientific literature and 
`r (trueCrowdData %>% filter(Species != "Unknown") %>% pull(Species) %>% unique() %>% length()) - 5` 
of which were novel species. All of the species identified in the video survey 
were cockatoos (_Cacatuidae_), Old World parrots (_Psittacinae_), or neotropical 
parrots (_Arinae_). The most common species in our survey, accounting for 
`r crowdData %>% filter(Species == "Pyrrhura molinae") %>% nrow()` videos from 
`r crowdData %>% filter(Species == "Pyrrhura molinae") %>% pull(IndividualID) %>% unique() %>% length()` 
individuals, was the green-cheeked conure (_Pyrrhura molinae_). In accordance 
with the scientific literature, the video survey found no evidence of tool use 
in any species of Psittaculidae, despite this family containing some of the most 
commonly kept pet species, including lovebirds, lorikeets, and Asian parakeet 
species. Combining both the video survey and the scientific literature, we can 
thus identify 28 total parrot species (7%) capable of true tool use as defined 
by Shumaker et al.[@Shumaker2011], compared to the 11 previously reported.

@@

Our data raise the question of why such a wide range of parrot species exhibit 
self-scratching tool use behaviours in captive environments. It is possible that 
these self-care behaviours emerge in captivity because pet parrots are kept with
few or no conspecifics, and therefore must innovate tool use to preen
inaccessible body areas, such as their heads and necks, which would have been
otherwise preened by flock members. This may indicate that human interaction and
allopreening in the absence of a conspecific is insufficient for pet parrots, as
suggested by studies on welfare and chronic stress in lone captive parrots 
[@Aydinonat2014;@Engebretson2006;@Mason2010;@Meehan2003]. Beyond social
isolation in captivity, other factors that reduce preening efficacy by an 
individual might also lead to the innovation of self-care tooling: for example,
a captive kea housed with conspecifics in a naturalistic environment was shown
to have innovated the use of a pebble tool for self-care likely as a consequence 
of a missing upper mandible [@Bastos2021]. These behaviours might therefore be 
innovated by individuals when ecologically necessary. Critically, however, a 
capacity to innovate a behaviour such as tool use must exist in a species in 
order to manifest in any environment, whether that be captive or in the wild. 
The fact that only a subset of parrot species are innovating self-care tooling 
in response to their captive setting and that self-care tooling is predicted by 
phylogenetic history and encephalisation strongly suggests that self-scratching 
indicates a capacity to produce tool use, which may be expressed in different 
ways in captive and wild settings. Interestingly, at present there is no
evidence to suggest that parrots exhibit self-scratching tool use in the wild.
This could reflect a true absence of this behaviour in wild parrots, which may
not experience the same environmental pressures that lead to the emergence of
self-scratching in captivity. On the other hand, it could also reflect a false
negative, with self-scratching occurring sufficiently infrequently or
inconspicuously in wild populations that it remains unobserved or unreported.

Our findings have a number of important implications. First, our findings
suggest that current research effort in the scientific literature is
insufficient to capture the real world occurrence of parrot tool use. If the
scientific literature had sampled the natural world sufficiently, we would
expect to see close correspondence between those species reported as tool users
in the literature and those species the public have filmed performing tool use. 
Instead, we discovered a large discrepancy between these two data sources, both 
in the prevalence of tool use and the species identified, likely due to the 
numerous difficulties inherent to conducting observations in the wild. This 
raises the possibility that other rare behaviours remain yet to be discovered in 
the scientific literature.

@@

Third, our results support existing theories of the drivers of tool use. We 
found that encephalisation was strongly positively related to the probability of 
tool use in our phylogenetic model, supporting previous theories linking 
relative brain size to increased tool innovation in birds [@Lefebvre1997;
@Lefebvre2002;@Lefebvre2004] and primates [@Reader2002]. We acknowledge that
encephalisation quotient is not a perfect measure due to measurement error and
challenges with interpretation [@Healy2007;@Logan2018]. However, encephalisation
quotient has much greater coverage across the parrot phylogeny than more 
fine-grained measures like whole neuron count [@Sol2022] and there is no reason 
to think that measurement error would produce the consistent patterns across our
study and prior work. To understand the causal mechanisms responsible for these
relationships, we encourage further work on the specific neural correlates of 
technical intelligence in parrots, e.g., Cabrera-Álvarez and Clayton
[@Alvarez2020]. In our phylogenetic model, we also found that tool use was 
somewhat more likely among feeding generalists compared to feeding specialists, 
although this difference was uncertain. This trend supports previous suggestions
that increased cognitive abilities and technical innovation rates are required
to expand a generalist species' dietary niche [@HenkeVonDerMalsburg2020;
@HenkeVonDerMalsburg2021;@Ducatez2015]. However, the trend contradicts theories
linking tool use to dietary specialisation, whereby species eating specific
foods that require extractive foraging have higher cognitive ability and are
especially prone to using tools [@HenkeVonDerMalsburg2020].

@@

## Limitations of study

A potential concern with data originating from digital video platforms is its 
reliability [@Auersperg2020;@DechaumeMoncharmont2020;@Farrar2020;@Nelson2013;@Sandor2020;@vonBayern2020]. 
In particular, the self-care tool use instances detected in our video survey 
might be attributable to training or unintentional cueing by the birds' owners.
However, there was little evidence to suggest that the observations of 
self-scratching tool use in the video survey were merely unintentional accidents 
or explicitly trained behaviours. Individual parrots often used tools slowly and 
repetitively over long periods of time, even across multiple different videos, 
suggesting that their behaviour was not random or accidental [@Sandor2020]. 
Parrots preferentially employed self-scratching tools on areas of their body 
that were otherwise inaccessible, with 96% of all instances involving scratching 
of the head or neck, suggesting intentional tool use. For 60% of the species in 
the video survey, we found two or more videos of repetitive and sustained 
scratching by different individuals of the same species in separate households, 
suggesting that the manipulations were intentional and recurring events that did 
not represent unusual stereotypies of any single individual. Regarding the 
possibility of training or cueing from owners, over half of the videos contained 
no evidence of human interaction aside from filming the behaviour. Humans only 
handed parrots their tools in two of the videos, and none of the videos featured 
owners directly rewarding tool use behaviours with food. It is unlikely that 
a range of owners would independently train parrots to scratch themselves with
tools. Finally, the high levels of phylogenetic signal in our data provide
strong evidence that the observations from our video survey reflect
biologically-endowed capacities for tool use rather than accidental or trained
behaviours, which would likely appear uniformly across the phylogeny. 
Nevertheless, future work using data such as this should choose target 
behaviours carefully and examine videos for evidence of training or cueing.

## Conclusion

@@

This project was made possible through the support of grants from the Templeton 
World Charity Foundation (A.H.T., X.J.N., TWCF-2018-0310, TWCF-2024-33225). The
authors would like to thank Daniel Sol for providing feedback on a previous
version of the manuscript.
